A dynamic bioenergetic model for coral-\emph{Symbiodinium} symbioses and coral bleaching as an alternate stable state

### 2017-07-11 ms/ms.tex
remake with revisions

Simulations using the discretized scheme were performed using R code
that is available in the data repository accompanying this article:
github.com/jrcunning/Rcoral. By experimentation, we found that a time
step of 0.1 days gave adequate precision for most simulations (including
used to generate Figs. 2-8 in this paper). For steady state estimations,
simulations were run until the change in specific growth rate of host
and symbiont and the S:H biomass ratio was less than 1e-5 per time step.
In regions of state space where very slow transient dynamics could be
expected (i.e.~near bifurcation points), sample steady state
calculations were verified using MATHEMATICA code for numerical root
finding (function FindRoot) with the code written independently by a
coauthor without reference to the R code.

@@

@@

The maximum predicted growth rates of \textasciitilde{}0.1
d\textsuperscript{-1}, occurring between \textasciitilde{}10-25 mol
photons m\textsuperscript{-2} s\textsuperscript{-1} light and
\textasciitilde{}4 µM DIN (Fig. 3A), are comparable to the rate of 0.07
d\textsuperscript{-1} measured by Tanaka et al. (2007) in \emph{Acropora
pulchra} under similar N-enriched conditions. Under conditions more
typical of reef environments (\textless{}0.5 µM DIN), predicted growth
rates are \textasciitilde{}0.01-0.03 d\textsuperscript{-1}. Observed
specific growth rates in several coral species fall near or below the
lower end of this range (\textasciitilde{}0.01 d\textsuperscript{-1})
(Osinga et al. 2011; Osinga et al. 2012), though values as high as 0.025
d\textsuperscript{-1} have been reported in \emph{Galaxea fascicularis}
(Schutter et al. 2010), and 0.04 d\textsuperscript{-1} in \emph{Aiptasia
diaphana}, a non-calcifying symbiotic anemone (Armoza-Zvuloni et al.
2014). However, it is not surprising that observed growth rates are
often lower than model predictions, since the model does not account for
ecological factors that may limit growth (e.g., competition, predation,
bioerosion). Furthermore, while most measurements are made on skeletal
growth, the model predicts biomass growth, which may not always be
strongly correlated (Anthony 2002).

@@

Since bleaching can be understood as a transition from a
nitrogen-limited state with high S:H biomass to a carbon-limited state
with low S:H biomass, induced by an external stressor, recovery can be
understood as a switch back to the nitrogen-limited state once the
external stressor is alleviated. In natural settings, this typically
occurs through seasonal declines in temperature and light. However,
hysteresis associated with the system's alternate stable states
indicates that the symbiosis cannot recover along the same trajectory it
followed during bleaching; indeed, the stressor must be alleviated
\emph{below} the threshold that initially caused bleaching in order for
the system to recover (Fig. 2). This is because under the same external
conditions, a bleached coral with low S:H biomass (relative to a healthy
coral with high S:H biomass) is characterized by greater light
amplification and weaker CCM activity, which enhance photooxidative
stress and serve to maintain the carbon-limited state. In order for the
system to recover, the stressor must be reduced enough such that
photooxidative stress ceases and translocation of carbon from symbiont
to host is resumed. Once this occurs, the host can energize its CCMs,
which further enhances carbon fixation and accelerates the system back
toward a nitrogen-limited state, indicative of recovery. The conditions
under which recovery can occur -- which determine the magnitude of
hysteresis (Fig. S9) -- depend on the relative abundance of nitrogen and
carbon in the environment. Higher food levels, representing a
non-autotrophic carbon source for the host, make it easier for the host
to overcome carbon-limitation, thus providing a potential mechanism
underlying observations that feeding aids recovery from bleaching
(Grottoli, Rodrigues, and Palardy 2006; Connolly, Lopez-Yglesias, and
Anthony 2012). Conversely, high external DIN impedes the
re-establishment of nitrogen-limitation, making recovery from bleaching
more difficult (i.e., narrowing the sets of conditions under which
recovery is possible).

Dynamic simulations of recovery reveal another interesting behavior of
the system: under some conditions, an `overshoot' occurs in which S:H
biomass temporarily increases beyond the ratio maintained in the
`healthy' state, before returning to stabilize at this value (Fig. 8).
In fact, unusually high symbiont densities have been observed following
bleaching in both experimental (Cunning, Silverstein, and Baker 2015)
and field studies (Kemp et al. 2014), and have been interpreted as a
potential `disequilibrium in host-symbiont regulation' (Kemp et al.
2014). The model reveals the dynamics of this `overshoot' as follows: as
symbionts repopulate the host, photosynthesis becomes increasingly
CO\textsubscript{2}-limited due to weak CCM activity of the
carbon-limited host. Thus, although symbiont growth is not yet
carbon-limited, a growing symbiont population has less and less excess
carbon (per symbiont) to share, and is thus moving toward
carbon-limitation. Meanwhile, because S:H biomass is increasing, the
host receives more and more carbon per unit host biomass, and is thus
moving away from carbon-limitation. If the host overcomes
carbon-limitation \emph{before} the symbiont reaches it, then the system
rapidly transitions to the nitrogen-limited state and the S:H ratio
stabilizes without an overshoot (Fig. 8A). However, if the symbiont
becomes carbon-limited first (Fig. 8B, 8C), then carbon translocation
per symbiont declines further and photosynthesis becomes more
CO\textsubscript{2}-limited, which maintains carbon-limitation of the
host. In this situation, continued growth of less and less productive
symbionts drives S:H biomass to a much higher level before the host
finally overcomes carbon-limitation. At this point, representing the
peak of the overshoot, the transition to nitrogen-limitation finally
takes place and the S:H biomass ratio declines and stabilizes as
positive growth is resumed.

Whether this `overshoot' occurs or not is determined by the relative
availability of carbon and nitrogen to the host -- any factor that
enhances carbon-limitation of host growth (e.g.~high DIN and/or low
feeding) therefore magnifies the overshoot and prolongs the
dysfunctional, carbon-limited state of the symbiosis (Fig. 8). On the
other hand, factors that favor nitrogen-limitation, such as low external
DIN and/or high feeding rates, will accelerate the re-establishment of
nitrogen-limitation and prevent an overshoot from occurring at all.
While many scenarios are possible under different environmental
conditions, we illustrate the general effects of varying N and C
availability on recovery from bleaching with a series of simulations
that vary the N:C ratio of host's heterotrophic food source (Fig. 8A-C):
lower N:C ratios (effectively representing lower DIN and/or higher
heterotrophy) favor nitrogen-limitation and more rapid recovery, while
higher N:C ratios (effectively representing higher DIN and/or lower
heterotrophy) favor carbon-limitation and prolonged recovery with a
larger overshoot in S:H biomass.

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig8.png}
\caption{Recovery from bleaching with varying N:C availability (A:
N:C=0.100; B: N:C=0.175; C: N:C=0.220). Higher N:C ratios in a
heterotrophic food source (effectively representing higher external DIN
and/or lower feeding rates) cause a larger overshoot in the S:H biomass
ratio, and prolong the duration of time until the system `recovers' by
re-establishing nitrogen-limitation. Substrate limitation coefficients
were calculated using Eq. 23. All simulations were run with default
parameters (except varying \(n_{NX}\)), with L=20 mol photons
m\textsuperscript{2} d\textsuperscript{-1}, DIN=1e-7 mol N
L\textsuperscript{-1}, prey=1e-7 CmolX L\textsuperscript{-1}, and
initial S:H biomass=0.001.}
\end{figure}

@@

This dynamic bioenergetic model of coral-\emph{Symbiodinium} symbioses
mechanistically reproduces patterns in steady-state coral growth and
symbiont abundance commonly observed in corals, including higher
symbiont abundance with higher nutrients and feeding, lower symbiont
abundance with increasing light, and optimal growth at intermediate
light levels. Moreover, the model reproduces complex dynamic behaviors
including seasonal changes in symbiont density at different nutrient
levels, rapid bleaching above a threshold of high light, mitigation of
bleaching by heterotrophic feeding, exacerbation of bleaching by
elevated nutrients, and an overshoot of symbiont density during recovery
from bleaching. These examples demonstrate the model's ability to
integrate multiple environmental forcing functions to reproduce complex
responses; meanwhile, the diversity of these phenomena suggest the model
has captured many of the important features of the system in a unifying
mechanistic framework. This model also provides a new conceptual
framework for considering coral bleaching as a transition to an
alternate stable state, which has important implications for
understanding the performance and maintenance of symbiotic interactions.
In this context, the `healthy' stable state represents a scenario in
which nitrogen-limitation stabilizes the symbiont to host biomass ratio
and maintains positive growth. Conversely, carbon-limitation represents
a dysfunctional state wherein positive feedbacks result in a loss of
photosynthetic capacity and negative growth. Importantly, these
`control' mechanisms arise entirely passively through feedbacks present
in the system.

In addition to the suite of specific examples presented here, the model
can be used to explore many different dynamic environmental scenarios,
and represents a tool that biologists and ecologists can use to generate
hypotheses and make predictions in both experimental and natural
settings. Moreover, parameter values can be modified to correspond to
different genetic or functional types of coral hosts and
\emph{Symbiodinium} partners in order to evaluate variability in system
responses. Thus, the diversity of potential applications for this model
is high, and we envision this work as a foundation for continued
development, which may include more detailed treatments of specific
modules (e.g., DIC processing), and the incorporation of more external
forcing capabilities (e.g., external DIC, temperature). Importantly,
open source R code allows this effort to benefit from contributions from
the wider scientific community, including those with empirical and
theoretical backgrounds. Ultimately, the continued refinement of these
tools is fundamental in elucidating the mechanisms of symbiosis function
and dysfunction, and in predicting coral responses to environmental
change.

@@

\hypertarget{ref-Cunning:2015ja}{}
Cunning, Ross, Rachel N Silverstein, and Andrew C Baker. 2015.
``Investigating the causes and consequences of symbiont shuffling in a
multi-partner reef coral symbiosis under environmental change.''
\emph{Proceedings of the Royal Society B}, 1--9.
doi:\href{https://doi.org/10.1098/rspb.2014.1725\%7B/\&\%7Ddomain=pdf\%7B/\&\%7Ddate_stamp}{10.1098/rspb.2014.1725\{\textbackslash{}\&\}domain=pdf\{\textbackslash{}\&\}date\_stamp}.

@@

>>>>

@@

Simulations using the discretized scheme were performed using R code
developed in the coRal R package \url{github.com/jrcunning/coRal}. By
experimentation, we found that a time step of 0.1 days gave adequate
precision for most simulations (including used to generate Figs. 2-8 in
this paper). For steady state estimations, simulations were run until
the changes in specific growth rate of the host and the S:H biomass
ratio were less than 1e-5 per time step. In regions of state space where
very slow transient dynamics could be expected (i.e.~near bifurcation
points), sample steady state calculations were verified using
MATHEMATICA code for numerical root finding (function FindRoot) with the
code written independently by a coauthor without reference to the R
code. All of the R code for the simulations and figures presented in
this paper can be found in the accompanying data repository at
\url{github.com/jrcunning/coRal-analysis}.

@@

The maximum predicted growth rates of \textasciitilde{}0.1
d\textsuperscript{-1}, occurring between \textasciitilde{}10-25 mol
photons m\textsuperscript{-2} d\textsuperscript{-1} light and
\textasciitilde{}4 µM DIN (Fig. 3A), are comparable to the rate of 0.07
d\textsuperscript{-1} measured by Tanaka et al. (2007) in \emph{Acropora
pulchra} under similar N-enriched conditions. Under conditions more
typical of reef environments (\textless{}0.5 µM DIN), predicted growth
rates are \textasciitilde{}0.01-0.03 d\textsuperscript{-1}. Observed
specific growth rates in several coral species fall near or below the
lower end of this range (\textasciitilde{}0.01 d\textsuperscript{-1})
(Osinga et al. 2011; Osinga et al. 2012), though values as high as 0.025
d\textsuperscript{-1} have been reported in \emph{Galaxea fascicularis}
(Schutter et al. 2010), and 0.04 d\textsuperscript{-1} in \emph{Aiptasia
diaphana}, a non-calcifying symbiotic anemone (Armoza-Zvuloni et al.
2014). However, it is not surprising that observed growth rates are
often lower than model predictions, since the model does not account for
ecological factors that may limit growth (e.g., competition, predation,
bioerosion). Furthermore, while most measurements are made on skeletal
growth, the model predicts biomass growth, which may not always be
strongly correlated (Anthony 2002).

@@

Since bleaching can be understood as a transition from a
nitrogen-limited state with high S:H biomass to a carbon-limited state
with low S:H biomass, induced by an external stressor, recovery can be
understood as a switch back to the nitrogen-limited state once the
external stressor is alleviated. In natural settings, this typically
occurs through seasonal declines in temperature and light. However,
hysteresis associated with the system's alternate stable states
indicates that the symbiosis cannot recover along the same trajectory it
followed during bleaching; indeed, the stressor must be alleviated
\emph{below} the threshold that initially caused bleaching in order for
the system to recover (Fig. 2). This is because under the same external
conditions, a bleached coral with low S:H biomass (relative to a healthy
coral with high S:H biomass) is characterized by greater light
amplification and weaker CCM activity, which enhance photooxidative
stress and serve to maintain the carbon-limited state. In order for the
system to recover, the stressor must be reduced enough such that
photooxidative stress ceases and translocation of carbon from symbiont
to host is resumed. Once this occurs, the host can energize its CCMs,
which further enhances carbon fixation and accelerates the system back
toward a nitrogen-limited state, indicative of recovery. The conditions
under which recovery can occur -- which determine the magnitude of
hysteresis (Fig. 8) -- depend on the relative abundance of nitrogen and
carbon in the environment. Higher food levels, representing a
non-autotrophic carbon source for the host, make it easier for the host
to overcome carbon-limitation (Fig. 8A, D), thus providing a potential
mechanism underlying observations that feeding aids recovery from
bleaching (Grottoli, Rodrigues, and Palardy 2006; Connolly,
Lopez-Yglesias, and Anthony 2012). Conversely, high external DIN impedes
the re-establishment of nitrogen-limitation, making recovery from
bleaching more difficult (i.e., narrowing - or eliminating - conditions
under which recovery is possible, Fig. 8C).

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig8.png}
\caption{Hysteresis under different nutrient and feeding regimes. Steady
state S:H biomass ratios in constant environments with (A) low DIN (1e-7
M) and low food (2e-7 M), (B) low DIN (1e-7 M) and no food, (C) high DIN
(2e-6 M) and low food (2e-7 M), and (D) high DIN (2e-6 M) and high food
(4e-7 M). In each panel, steady states are shown starting from both high
initial S:H biomass (1, i.e.~healthy corals - closed circles) and low
initial initial S:H biomass (0.0001, i.e.~bleached corals - open
circles). Points are colored according to whether the host exhibits
positive (black) or negative (red) growth at steady state.}
\end{figure}

Dynamic simulations of recovery reveal another interesting behavior of
the system: under some conditions, an `overshoot' occurs in which S:H
biomass temporarily increases beyond the ratio maintained in the
`healthy' state, before returning to stabilize at this value (Fig. 9).
In fact, unusually high symbiont densities have been observed following
bleaching in both experimental (Cunning, Silverstein, and Baker 2015)
and field studies (Kemp et al. 2014), and have been interpreted as a
potential `disequilibrium in host-symbiont regulation' (Kemp et al.
2014). The model reveals the dynamics of this `overshoot' as follows: as
symbionts repopulate the host, photosynthesis becomes increasingly
CO\textsubscript{2}-limited due to weak CCM activity of the
carbon-limited host. Thus, although symbiont growth is not yet
carbon-limited, a growing symbiont population has less and less excess
carbon (per symbiont) to share, and is thus moving toward
carbon-limitation. Meanwhile, because S:H biomass is increasing, the
host receives more and more carbon per unit host biomass, and is thus
moving away from carbon-limitation. If the host overcomes
carbon-limitation \emph{before} the symbiont reaches it, then the system
rapidly transitions to the nitrogen-limited state and the S:H ratio
stabilizes without an overshoot (Fig. 9A). However, if the symbiont
becomes carbon-limited first (Fig. 9B, 9C), then carbon translocation
per symbiont declines further and photosynthesis becomes more
CO\textsubscript{2}-limited, which maintains carbon-limitation of the
host. In this situation, continued growth of less and less productive
symbionts drives S:H biomass to a much higher level before the host
finally overcomes carbon-limitation. At this point, representing the
peak of the overshoot, the transition to nitrogen-limitation finally
takes place and the S:H biomass ratio declines and stabilizes as
positive growth is resumed.

Whether this `overshoot' occurs or not is determined by the relative
availability of carbon and nitrogen to the host -- any factor that
enhances carbon-limitation of host growth (e.g.~high DIN and/or low
feeding) therefore magnifies the overshoot and prolongs the
dysfunctional, carbon-limited state of the symbiosis (Fig. 9). On the
other hand, factors that favor nitrogen-limitation, such as low external
DIN and/or high feeding rates, will accelerate the re-establishment of
nitrogen-limitation and prevent an overshoot from occurring at all.
While many scenarios are possible under different environmental
conditions, we illustrate the general effects of varying N and C
availability on recovery from bleaching with a series of simulations
that vary the N:C ratio of host's heterotrophic food source (Fig. 9A-C):
lower N:C ratios (effectively representing lower DIN and/or higher
heterotrophy) favor nitrogen-limitation and more rapid recovery, while
higher N:C ratios (effectively representing higher DIN and/or lower
heterotrophy) favor carbon-limitation and prolonged recovery with a
larger overshoot in S:H biomass.

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig9.png}
\caption{Recovery from bleaching with varying N:C availability (A:
N:C=0.100; B: N:C=0.175; C: N:C=0.220). Higher N:C ratios in a
heterotrophic food source (effectively representing higher external DIN
and/or lower feeding rates) cause a larger overshoot in the S:H biomass
ratio, and prolong the duration of time until the system `recovers' by
re-establishing nitrogen-limitation. Substrate limitation coefficients
were calculated using Eq. 23. All simulations were run with default
parameters (except varying \(n_{NX}\)), with L=20 mol photons
m\textsuperscript{2} d\textsuperscript{-1}, DIN=1e-7 mol N
L\textsuperscript{-1}, prey=1e-7 CmolX L\textsuperscript{-1}, and
initial S:H biomass=0.001.}
\end{figure}

@@

This dynamic bioenergetic model of coral-\emph{Symbiodinium} symbioses
mechanistically reproduces patterns in steady-state coral growth and
symbiont abundance commonly observed in corals, including higher
symbiont abundance with higher nutrients and feeding, lower symbiont
abundance with increasing light, and optimal growth at intermediate
light levels. Moreover, the model reproduces complex dynamic behaviors
including seasonal changes in symbiont density at different nutrient
levels, rapid bleaching above a threshold of high light, mitigation of
bleaching by heterotrophic feeding, exacerbation of bleaching by
elevated nutrients, and an overshoot of symbiont density during recovery
from bleaching. These examples demonstrate the model's ability to
integrate multiple environmental forcing functions to reproduce complex
responses; meanwhile, the diversity of these phenomena suggest the model
has captured many of the important features of the system in a unifying
mechanistic framework. This model also provides a new conceptual
framework for considering coral bleaching as a transition to an
alternate stable state, which has important implications for
understanding the performance and maintenance of symbiotic interactions.
In this context, the `healthy' stable state represents a scenario in
which nitrogen-limitation stabilizes the symbiont to host biomass ratio
and maintains positive growth. Conversely, carbon-limitation represents
a dysfunctional state wherein positive feedbacks result in coral
bleaching and negative growth, with hysteresis maintaining this
alternate state until a favorable environment is re-established.
Interestingly, alternate stable states have received much attention in
coral reef community ecology with respect to coral- and
macroalgal-dominance on reefs (Mumby, Hastings, and Edwards 2007). Here,
we find that the same principles may also mediate bleaching and recovery
in individual corals.

The model developed here can be used to explore many different dynamic
environmental scenarios, and represents a tool that biologists and
ecologists can use to generate hypotheses and make predictions in both
experimental and natural settings. Moreover, parameter values can be
modified to correspond to different genetic or functional types of coral
hosts and \emph{Symbiodinium} partners in order to evaluate variability
in system responses. Thus, the diversity of potential applications for
this model is high, and we envision this work as a foundation for
continued development, which may include more detailed treatments of
specific modules (e.g., DIC processing), and the incorporation of more
external forcing capabilities (e.g., external DIC, temperature).
Importantly, open source R code allows this effort to benefit from
contributions from the wider scientific community, including those with
empirical and theoretical backgrounds. Ultimately, the continued
refinement of these tools is fundamental in elucidating the mechanisms
of symbiosis function and dysfunction, and in predicting coral responses
to environmental change.

\section{Acknowledgements}\label{acknowledgements}

This work was supported by a NSF Postdoctoral Research Fellowship in
Biology to RC (\#1400787).

@@

\hypertarget{ref-Cunning:2015ja}{}
Cunning, Ross, Rachel N Silverstein, and Andrew C Baker. 2015.
``Investigating the causes and consequences of symbiont shuffling in a
multi-partner reef coral symbiosis under environmental change.''
\emph{Proceedings of the Royal Society B} 282: 20141725.
doi:\href{https://doi.org/10.1098/rspb.2014.1725\%7B/\&\%7Ddomain=pdf\%7B/\&\%7Ddate_stamp}{10.1098/rspb.2014.1725\{\textbackslash{}\&\}domain=pdf\{\textbackslash{}\&\}date\_stamp}.

@@

\hypertarget{ref-Mumby:2007cr}{}
Mumby, Peter J, Alan Hastings, and Helen J. Edwards. 2007. ``Thresholds
and the resilience of Caribbean coral reefs.'' \emph{Nature} 450 (7166).
Nature Publishing Group: 98--101.
doi:\href{https://doi.org/10.1038/nature06252}{10.1038/nature06252}.
